\thetitle

### 2021-03-01 writing/introduction.tex
intro, landscape, why-rational: typo, wording, and other small fixes

This paper's {\em first contribution\/} is {\em a method for evaluating the
effectiveness of blame assignment strategies\/} in the gradual typing world.
The top-level innovation is the idea of a {\em rational programmer\/}, that is,
a programmer that acts only in response to available information (see
sec.~\ref{sec:why-rational}). In the case of an impedance mismatch, the
available information consists of the error message and the current state of the
program. The rational programmer can hence use the former to change the
latter---and this systematic, information-driven process can be
implemented.\footnote{\citet{cc-snapl-19} use this tracing method to localize
type errors in inference-based gradual typing systems.}  Implementing it means
overcoming major challenges: injecting representative impedance mismatches;
putting the various kinds of error information to comparable use; and sampling
the huge space of possibilities (see sec.~\ref{sec:challenges}).

The paper's {\it second contribution\/} is a set of {\em results from applying
the evaluation method to three distinct blame assignment strategies and
approximately 72,200 cases\/} (see sec.~\ref{sec:results}): (1) {\it
Transient\/}, i.e. Reticulated's tracking of typed/untyped boundaries; (2) {\it
Natural\/}, i.e. the use of higher-order contract system and its blame
assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e. the approach of
industrial systems, which forgo blame in favor of error messages from the safety
checks of the underlying language.---The results (see
sec.~\ref{sec:discussion}) are at least somewhat surprising.  In principle
they validate the conjectures behind the work of theoreticians.  First, a good
blame assignment strategy helps with the search for impedance mismatches between
the specified types and the behavior of untyped components.  Second, Natural's
wrapper-based blame tracking is more useful than Transient's ``collective
blame'' tracking algorithm, which in turn is superior to Erasure. {\em But\/},
the application of the method also indicates problems with the expectations of
theoreticians. The first problem is that the cost of Transient's blame can be
huge. The second one concerns the difference between the methods; neither is
{\it Natural\/} vastly better than {\it Transient\/} nor are the two blame
assignment methods clearly preferable to {\it Erasure\/}.  In turn, these
problems suggests that, on one hand, the existing theory does not predict
practice properly, and on the other hand, the existing practice may need to find
additional ways to collect data.

>>>>

This paper's {\em first contribution\/} is {\em a method for evaluating the
effectiveness of blame assignment strategies\/} in the gradual typing world.
The top-level innovation is the idea of a {\em rational programmer\/}, that is,
a programmer that acts only in response to available information (see
sec.~\ref{sec:why-rational}). In the case of an impedance mismatch, the
available information consists of the error message and the current state of the
program. The rational programmer can hence use the former to change the
latter---and this systematic, information-driven process can be
implemented and tested at scale on real programs.\footnote{\citet{cc-snapl-19} use this tracing method to localize
type errors in inference-based gradual typing systems.}  Testing it systematically requires
overcoming major challenges: injecting representative impedance mismatches;
putting the various kinds of error information to comparable use; and sampling
the huge space of possibilities (see sec.~\ref{sec:challenges}).

The paper's {\it second contribution\/} is a set of {\em results from applying
the evaluation method to three distinct blame assignment strategies in
approximately 72,200 different scenarios\/} (see sec.~\ref{sec:results}): (1)~{\it
Transient\/}, i.e. Reticulated's tracking of typed/untyped boundaries~\cite{vss-popl-2017}; (2)~{\it
Natural\/}, i.e. the use of a higher-order contract system and its blame
assignment~\cite{ff-icfp-2002}; and (3) {\it Erasure\/}, i.e. the approach of
industrial systems, which forgo blame in favor of error messages from the safety
checks of the underlying language. The results (see
sec.~\ref{sec:discussion}) are at least somewhat surprising.  In principle
they validate the conjectures behind the work of theoreticians.  First, a good
blame assignment strategy helps with the search for impedance mismatches between
the specified types and the behavior of untyped components.  Second, Natural's
wrapper-based blame tracking is more useful than Transient's ``collective
blame'' tracking algorithm, which in turn is superior to Erasure. {\em But\/},
the application of the method also indicates problems with the expectations of
theoreticians. The first problem is that the cost of Transient's blame can be
huge. The second one concerns the difference between the methods; neither is
{\it Natural\/} vastly better than {\it Transient\/} nor are the two blame
assignment methods clearly preferable to {\it Erasure\/}.  In turn, these
problems suggests that, on one hand, the existing theory does not predict
practice properly, and on the other hand, the existing practice may need to find
additional ways to collect data.

### 2021-03-01 writing/landscape.tex
justify rational programmer section
intro, landscape, why-rational: typo, wording, and other small fixes

\item[\texttt{crypto-pkac-lib}] (at the bottom left) also imports \texttt{pack}
and assumes for it the same type as {\tt typed-pack-lib}. It applies the
function in the definition of the exported {\tt crypto-pack} function, which
encrypts its input before passing it to \texttt{pack}.

@@

Under {\it the Natural semantics\/}, functions imported into and exported from
typed modules are wrapped in proxies that enforce the static type discipline
with run-time checks and track responsibilities~\cite{tf-popl-2008,
tfffgksst-snapl-2017}. Thus, when {\tt pack} is imported into a typed module,
the run-time system checks that it is a function and wraps it in a protective
proxy, which in turn, enforces the type of the function with run-time checks.
In essence, this check protects typed modules when they call an untyped function
because it may not produce the promised kind of value, as is the case here with
{\tt pack} from {\tt pack-lib}. Analogously the run-time system wraps exported
functions such as {\tt crypto-pack} of a typed modules in a proxy that checks
its arguments.  This check protects functions exported from type modules against
applications in untyped code that supply the wrong kind of argument value.

As this analysis implies, if the return-type check fails, the problem is with
the untyped module that supplied an incorrect function, here {\tt pack-lib}. If
the argument-type check fails, responsibility lies with {\tt client}. Due to the
proxies wrapped around functions, Natural can easily track the responsible party
for each check. Thus, in the example of figure~\ref{fig:example}, as
\texttt{typed-pack} returns, the return-type check fails, and Natural blames
\texttt{typed-pack-lib}.

Under {\it the Transient semantics\/}, typed code is compiled so that all entry
points to functions check their arguments at run time and all function calls
check their return values against the expected type~\cite{vss-popl-2017}.
Furthermore, Transient uses \emph{shallow} checks, meaning they inspect only the
top constructor of a value. Since retrieving a value from within a structure (or
list, array, hash table etc.) is performed via a function call, the innards of a
complex value are checked on a piecemeal basis.

As a result, the call to \texttt{typed-pack} does {\em not\/} signal an error
because it takes place in the untyped {\tt client} module, which is compiled in
the usual manner. Because {\tt pack} is called in the {\tt crypto-pack-lib}
module, Transient's inlined checks make sure that the imported
\texttt{pack} is a function and that the result of its call in is a list. This
last check fails when \texttt{client} calls to {\tt crypto-pack}.

In order to pinpoint the responsible party for failed checks, Transient
maintains a map from values to the boundaries between typed/untyped modules that
they cross, plus the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{pack-lib} to {\tt crypto-pack-lib} and
from {\tt crypto-pack-lib} to {\tt client} with the type that appears in the
{\tt required/typed} forms in the example. Since the failed check corresponds to
the return type of \texttt{pack}, Transient blames the sources of the two
boundary crossings, which in this particular case is \texttt{pack-lib}. In
general though, Transient blames more than one party. In fact, the theoretical
work of~\citet{gfd-oopsla-2019} shows that for some programs Transient
constructs a blame set that excludes responsible parties and includes modules
irrelevant to the failing check.

@@

The following table summarizes the illustration:
%% -----------------------------------------------------------------------------
\begin{center}
\begin{tabular}{l|ll}
                                 \multicolumn{3}{c}{the evaluation of} \\
\relax                  & {\tt public-data}             & {\tt secret-data} in {\tt client}, \\
{under the $\cdot$ semantics yields} & & \\ \hline
%% ----------------------------------------------------------------------------------------
\qquad {\it Natural\/}   & error, blaming              & error, blaming \\
     		   	 & \quad {\tt typed-pack-lib}  & \quad {\tt typed-pack-lib} \\
\qquad {\it Transient\/} & no error   		       & error, blaming the {\tt pack-lib}/ \\
      		  	 &    			       & \quad {\tt crypto-pack-lib} boundary \\
\qquad {\it Erasure\/}   & no error 		       & no error \\
\end{tabular}
\end{center}
%% -----------------------------------------------------------------------------

>>>>

\item[\texttt{crypto-pack-lib}] (at the bottom left) also imports \texttt{pack}
and assumes for it the same type as {\tt typed-pack-lib}. It applies the
function in the definition of the exported {\tt crypto-pack} function, which
encrypts its input before passing it to \texttt{pack}.

@@

Under {\it the Natural semantics\/}, functions imported into and exported from
typed modules are wrapped in proxies that enforce the static type discipline
with run-time checks and track responsibilities~\cite{tf-popl-2008,
tfffgksst-snapl-2017}. Thus, when {\tt pack} is imported into a typed module,
the run-time system checks that it is a function and wraps it in a protective
proxy, which in turn, enforces the type of the function with run-time checks.
In essence, this check protects typed modules when they call an untyped function
because it may not produce the promised kind of value, as is the case here with
{\tt pack} from {\tt pack-lib}. Analogously the run-time system wraps each exported
function of a typed module (such as {\tt crypto-pack}) in a proxy that checks
its arguments.  This check protects functions exported from typed modules against
applications in untyped code that supply the wrong kind of argument value.

As this analysis implies, if a return-type check fails, the problem is with
the untyped module that supplied an incorrect function, here {\tt pack-lib}. If
an argument-type check fails, responsibility lies with {\tt client}. Due to the
proxies wrapped around functions, Natural can easily track the responsible party
for each check. Thus, in the example of figure~\ref{fig:example}, as
\texttt{pack} returns, the return-type check fails, and Natural blames
\texttt{pack-lib}.

Under {\it the Transient semantics\/}, typed code is compiled so that all entry
points to functions check their arguments at run time and all function calls
check their return values against the expected type~\cite{vss-popl-2017}.
Furthermore, Transient uses \emph{shallow} checks, meaning they inspect only the
top constructor of a value. Since retrieving a value from within a structure (or
list, array, hash table etc.) is performed via a function call, the contents of a
complex value are checked on a piecemeal basis.

As a result, the call to \texttt{typed-pack} does {\em not\/} signal an error
because it takes place in the untyped {\tt client} module, which is compiled in
the usual manner. Later, when {\tt pack} is called in the {\tt crypto-pack-lib}
module, Transient's inlined checks make sure that the imported
\texttt{pack} is a function and that its result is a list. This
last check fails in \texttt{client}'s call to {\tt crypto-pack}.

In order to locate the responsible party for failed checks, Transient
maintains a map from values to the boundaries between typed/untyped modules that
they cross, plus the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{pack-lib} to {\tt crypto-pack-lib} and
from {\tt crypto-pack-lib} to {\tt client} with the type that appears in the
{\tt required/typed} forms in the example. Since the failed check corresponds to
the return type of \texttt{pack}, Transient blames the sources of the two
boundary crossings, which in this particular case is \texttt{pack-lib}. In
general though, Transient blames more than one party. In fact, the theoretical
work of~\citet{gfd-oopsla-2019} shows that for some programs Transient
constructs a blame set that excludes responsible parties and includes modules
irrelevant to the failing check.

@@

The following table summarizes the illustration:
%% -----------------------------------------------------------------------------
\begin{center}
\begin{tabular}{l|ll}
                        &        \multicolumn{2}{|c}{the evaluation of} \\
\relax                  & {\tt public-data}               & {\tt secret-data} in {\tt client}, \\
{under the $\cdot$ semantics yields} & & \\ \hline
%% ----------------------------------------------------------------------------------------
\qquad {\it Natural\/}   & error, blaming {\tt pack-lib}  & error, blaming {\tt pack-lib} \\
     		   	 &                                & \\
\qquad {\it Transient\/} & no error   		          & error, blaming the boundary \\
      		  	 &    			          & \quad {\tt pack-lib}/{\tt crypto-pack-lib} \\
\qquad {\it Erasure\/}   & no error 		          & no error \\
\end{tabular}
\end{center}
%% -----------------------------------------------------------------------------

### 2021-03-01 writing/why-rational.tex
intro, landscape, why-rational: typo, wording, and other small fixes

This paper dubs Lazarek et. alii's idea {\em rational programmer\/}. Like Stuart
Mill's {\it homo economicus\/}, the rational programmer approximates the
behavior of a software developer who reduces time spent on a task given the
available information. In the context of gradual typing, the rational programmer
has two pieces of information when an impedance mismatch signals exceptional
behavior: the error message and the state of the program. Hence, the most
rational procedure is to use the former to improve the latter. Put differently,
a rational programmer translates the Wadler-Findler slogan into a debugging
method, searching for the source of the impedance mismatch with additional type
annotations. Measuring this simulated programmer's behavior on a large number of
debugging scenarios yields data that is similar to data collected in an
empirical, human-facing manner.

@@

Now consider the concrete (and simplistic) example of figure~\ref{fig:rational}.
Initially the code base consists of the two \dyncolor\ module on the left plus
the \typecolor\ module at the bottom; \dyncolor\ indicates untyped, while
\typecolor\ means typed. When a typed module imports an untyped module, it must
assign types to the imported identifiers for the type checker's sake. Here {\tt
main} specifies that {\tt na-client} consumes a {\tt Real} number and produces a
non-positive one.\footnote{Typed Racket's type system reflects Racket's subset
reasoning about numbers, not (only) machine-level
representations~\cite{stathff-padl-12}.} A program execution ends in this error:
\begin{verbatim}
    na-client: broke its own contract
      promised: (<=/c 0)
      produced: 40
      in: (-> any/c (<=/c 0))           
      contract from: (interface for na-client)
      blaming: (interface for na-client)
       (assuming the contract is correct)
\end{verbatim}
The referenced contract is the compilation of the type of {\tt na-client}. The
definitive hint is ``blaming: (interface for na-client)'' with the caveat
``(assuming the contract is correct).''

Assuming the rational programmer trust the type of {\tt na-client}, the next
step is to inspect the {\tt layer} module and to equip it with type
annotations. The result is the \typecolor\ module in the middle, and {\tt
main}'s import is now re-directed there: {\tt (submod ".." layer)}. Running the
program again yields this message:
\begin{verbatim}
    neg-abs: broke its own contract
      promised: (<=/c 0)
      produced: 10
      in: (-> any/c (<=/c 0))
      contract from: (interface for neg-abs)
      blaming: (interface for neg-abs)
       (assuming the contract is correct)
\end{verbatim}
The last step of the rational programmer is to assign types to the {\tt server}
module. At this point the type checker objects to the conjecture type of {\tt
neg-abs}---and the source of the impedance mismatch is found. How to fix it is a
separate question.

Like {\it homo economicus\/}, the rational programmer is an
approximation.  People don't behave purely rationally as economic
actors, and they also don't do so as software developers. The point is
not to deny the existence of ``lucky hunches'' programmers or
``tinkering works'' approaches and so on. It is also not to deny that
equipping entire modules with types represents an always feasible
approach.  But, the concept of studying the idea of an economically
rational actor has produced benefits to political economics, and this
paper suggests that implementing and studying the rational programmer
will help language designers.

>>>>

This paper dubs Lazarek et. all's idea {\em the rational programmer\/}. Like Stuart
Mill's {\it homo economicus\/}, the rational programmer approximates the
behavior of a software developer who reduces time spent on a task given the
available information. In the context of gradual typing, the rational programmer
has two pieces of information when an impedance mismatch signals exceptional
behavior: the error message and the state of the program. Hence, the most
rational procedure is to use the former to improve the latter. Specifically,
a rational programmer translates the Wadler-Findler slogan into a debugging
method, searching for the source of the impedance mismatch by adding type
annotations to the parts of the program identified in the error.
If the type checker rejects an annotation, the rational programmer has found the problem.
Measuring this simulated programmer's behavior on a large number of
debugging scenarios yields data that is similar to data collected in an
empirical, human-facing manner.

@@

Now consider the concrete (and simplistic) example of figure~\ref{fig:rational}.
Initially the code base consists of the two \dyncolor\ modules on the left plus
the \typecolor\ module at the bottom; \dyncolor\ indicates untyped, while
\typecolor\ means typed. When a typed module imports an untyped module, it must
assign types to the imported identifiers for the type checker's sake. Here {\tt
main} specifies that {\tt na-client} consumes a {\tt Real} number and produces a
non-positive one.\footnote{Typed Racket's type system reflects Racket's subset
reasoning about numbers, not (only) machine-level
representations~\cite{stathff-padl-12}.} A program execution ends in this error:
\begin{verbatim}
    na-client: broke its own contract
      promised: (<=/c 0)
      produced: 40
      in: (-> any/c (<=/c 0))           
      contract from: (interface for na-client)
      blaming: (interface for na-client)
       (assuming the contract is correct)
\end{verbatim}
The referenced contract is the compilation of the type of {\tt na-client}. The
definitive hint is ``{\tt blaming: (interface for na-client)}'' with the caveat
``{\tt (assuming the contract is correct)}.''

Assuming the rational programmer trusts the type of {\tt na-client}, the next
step is to inspect the {\tt layer} module and to equip it with type
annotations. This results in the \typecolor\ module in the middle, and {\tt
main}'s import is now re-directed there by {\tt (submod ".." layer)}. Running the
program again yields a new message:
\begin{verbatim}
    neg-abs: broke its own contract
      promised: (<=/c 0)
      produced: 10
      in: (-> any/c (<=/c 0))
      contract from: (interface for neg-abs)
      blaming: (interface for neg-abs)
       (assuming the contract is correct)
\end{verbatim}
The last step of the rational programmer is to assign types to the {\tt server}
module. At this point the type checker objects to the conjecture type of {\tt
neg-abs}---and the source of the impedance mismatch is found. How to fix it is a
separate question.

Like {\it homo economicus\/}, the rational programmer is an
approximation.  People don't behave purely rationally as economic
actors, and they also don't do so as software developers. The point is
not to deny the existence of ``lucky hunches'' programmers or
``tinkering works'' approaches and so on. It is also not to deny that
equipping entire modules with types represents an always feasible
approach.  But the concept of studying the idea of an economically
rational actor has produced benefits to political economics, and this
paper suggests that implementing and studying the rational programmer
will help language designers.
